Methods for Analyzing Models of\\ Rod-Shaped Bacteria

### 2025-11-30 paper/main.tex
rename section for imaging-metric
rename subsection

\subsubsection{Direct Skeleton-Space Metric}
\label{subsec:skeleton-metric}
In the scenario where no cells divide during the observed interval, we can use our extraction
algorithm to obtain the position of the cells and then compare them directly to the cellular
skeletons obtained from the computational results.
Given $N_c$ cells with $N_v$ vertices in $d$ dimensions at $N_t$ time points, the total information
can be stored in an array $\nu_{i,j,k,l}$ of shape $(N_c,N_v,d,N_t)$.
Comparing the results of a simulation $\nu$ with experimental values $\tilde{\nu}$ is thus only a
matter of comparing two of these arrays via a suitable cost function~\cite{Wang2020}.
In our approach we utilize a \ac{ls} approach given by

@@

\subsubsection{Imaging-Space Metric}
\label{subsec:image-space-metric}
\paragraph{Motivation}
The preceding algorithm is performant and effective but cannot be applied in the case when a
division event splits apart the bacterium since this would require comparing sets of unequal
cardinality.
This situation is displayed in figure~\ref{fig:mask-difference-metric} (E) where two distinct
bacterial configurations originate from the same initial state but contain varying numbers of
bacteria at the next time step.
Such a mismatch can happen, e.g. due to non-matching division times of model and data.
Thus a cell which is present in the simulation might not yet have a corresponding cell in the
experimental data at this time point yet (or vice-versa).

@@

We extract the positions of the rod from the data images (see
Figure~\ref{fig:amir-bending-simulation} (A)) and compare the simulated positions with the extracted
ones using the direct comparison algorithm (see Subsection~\ref{subsec:skeleton-metric}).
Our simulation is split into two phases; One with the external flow force that initially bends the
rod and lasts $\SI{7}{\second}$ and a second phase during which the force is disabled and the cell
returns to its original shape over a span of $\SI{17}{\second}$.

@@

In this section we study a collection of growing, dividing and physically interacting
\ac{ecoli}.
We use a series of microscopic images and corresponding segmentation masks for the individual
bacteria as the underlying dataset which is partially displayed in
Figure~\ref{fig:application-2-data-overview}.
With the previously developed extraction algorithm (see
section~\ref{subsec:position-extraction-algorithm}) we obtain the positions of our agents at all
time points.
These can be used as initial values and later within our parameter estimation routine.
We initialize our computational model (see section~\ref{sec:computational-model}) with the obtained
positions and optimize the parameters in order to find a set which best describes the observed
results.
The first case considers a time interval without cell-division while the second case does contain
division events.
Therefore, the first approach will optimize the direct comparison cost function (see
section~\ref{subsec:skeleton-metric}) while the latter approach has to account for division events
and thus requires the image-space metric method (see section~\ref{subsec:image-space-metric}).
These two scenarios present an opportunity to apply the previously developed methods within a
realistic context and validate our approach.

@@

In this section we will focus on a time-interval which does not contain any division events.
By doing so, we can utilize the simpler and more performant direct comparison scheme (see
\ref{subsec:skeleton-metric}) for our cost function.
We have chosen said interval such that effects related to the lag phase can be neglected which
allows us to sample a lower number of paremeters thus improving runtime performance of our
optimization scheme.

>>>>

\subsubsection{Direct Position-Space Metric}
\label{subsec:direct-metric}
In the scenario where no cells divide during the observed interval, we can use our extraction
algorithm to obtain the position of the cells and then compare them directly to the cellular
skeletons obtained from the computational results.
Given $N_c$ cells with $N_v$ vertices in $d$ dimensions at $N_t$ time points, the total information
can be stored in an array $\nu_{i,j,k,l}$ of shape $(N_c,N_v,d,N_t)$.
Comparing the results of a simulation $\nu$ with experimental values $\tilde{\nu}$ is thus only a
matter of comparing two of these arrays via a suitable cost function~\cite{Wang2020}.
In our approach we utilize a \ac{ls} approach given by

@@

\subsubsection{Lineage-aware Imaging-Space Metric}
\label{subsec:image-space-metric}
\paragraph{Motivation}
The preceding algorithm is performant and effective but cannot be applied in the case when a
division event splits apart the bacterium since this would require comparing sets of unequal
cardinality.
This situation is displayed in figure~\ref{fig:mask-difference-metric} (E) where two distinct
bacterial configurations originate from the same initial state but contain varying numbers of
bacteria at the next time step.
Such a mismatch can happen, e.g. due to non-matching division times of model and data.
Thus a cell which is present in the simulation might not yet have a corresponding cell in the
experimental data at this time point yet (or vice-versa).

@@

We extract the positions of the rod from the data images (see
Figure~\ref{fig:amir-bending-simulation} (A)) and compare the simulated positions with the extracted
ones using the direct comparison algorithm (see Subsection~\ref{subsec:direct-metric}).
Our simulation is split into two phases; One with the external flow force that initially bends the
rod and lasts $\SI{7}{\second}$ and a second phase during which the force is disabled and the cell
returns to its original shape over a span of $\SI{17}{\second}$.

@@

In this section we study a collection of growing, dividing and physically interacting
\ac{ecoli}.
We use a series of microscopic images and corresponding segmentation masks for the individual
bacteria as the underlying dataset which is partially displayed in
Figure~\ref{fig:application-2-data-overview}.
With the previously developed extraction algorithm (see
section~\ref{subsec:position-extraction-algorithm}) we obtain the positions of our agents at all
time points.
These can be used as initial values and later within our parameter estimation routine.
We initialize our computational model (see section~\ref{sec:computational-model}) with the obtained
positions and optimize the parameters in order to find a set which best describes the observed
results.
The first case considers a time interval without cell-division while the second case does contain
division events.
Therefore, the first approach will optimize the direct comparison cost function (see
section~\ref{subsec:direct-metric}) while the latter approach has to account for division events
and thus requires the image-space metric method (see section~\ref{subsec:image-space-metric}).
These two scenarios present an opportunity to apply the previously developed methods within a
realistic context and validate our approach.

@@

In this section we will focus on a time-interval which does not contain any division events.
By doing so, we can utilize the simpler and more performant direct comparison scheme (see
\ref{subsec:direct-metric}) for our cost function.
We have chosen said interval such that effects related to the lag phase can be neglected which
allows us to sample a lower number of paremeters thus improving runtime performance of our
optimization scheme.
